\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-13 tex/paper.tex
sentence for Hamiltonian ABC.. lacking context.

\begin{abstract}
Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. In this note, we develop Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model to
continuous or discrete  data. We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

@@

Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation. In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters that are endowed with some level of interpretation. In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness. 
Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
be a fixed-size vector with a simple density. Instead, the dimensionality of $z$ and the nature of it's components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way. Moreover, the dimensionality of $\bfz$ may be much larger than the dimensionality of $\bfx$.

@@

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$. Operationally, %That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the density $p(\mathbf{x}|\bftheta)$ can be 
rewritten as
%considered as a non-differentiable reparameterization of the random variable $\bfz$
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.
%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
%used to generate data in forward mode. For this reason, evaluating the integral
%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
%e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters 
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence. For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate.

@@

\subsection{Adversarial variational learning}

\kcnote{connect to MWE and MEWE here?}

In order to target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially}, Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.  Each of these assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model. 
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of a rejection sampler in an ABC-like setup (as opposed to variational inference).  They emphasize that this approach allows one to eliminate the summary statistics typically used in ABC and calculate the Wasserstein distance explicitly without use Kantorovich-Rubinstein duality and the critic $d$. For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection of the data onto Hilbert space-filling curves. Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality. They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' \kcnote{Can we evaluate their Eq. 6 with our Eq. 9} Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$. This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.
\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL. We also don't specify a prior b/c it's like implicit empirical bayes. They also don't specify a recognition model to latents. }

@@

As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
intimately tied to a class of algorithms that can be framed as density
estimation-by-comparison. In most cases, these inference algorithms are
formulated as an iterative two-step process where the model distribution is
first compared to the true data distribution and then updated to make it more
comparable to the latter. Relevant work in this direction includes \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood}.

@@

In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients ... \kcnote{todo}

>>>>

\begin{abstract}
Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model to
continuous or discrete  data. \kcnote{emperical variational bayes}. We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

@@

Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation. In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters and they are endowed with some level of interpretation. In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness. 
Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
be a fixed-size vector with a simple density. Instead, the dimensionality of $\bfz$ and the nature of its components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way. Moreover, the dimensionality of $\bfz$ may be much larger than the dimensionality of $\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.

@@

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$. Operationally, %That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the density $p(\mathbf{x}|\bftheta)$ can be 
written as
%considered as a non-differentiable reparameterization of the random variable $\bfz$
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.
%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
%used to generate data in forward mode. For this reason, evaluating the integral
%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
%e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters 
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.
%~\citep{bernton2017inference}. 
%For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate if $ \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta^*)) \to 0$.

%By plugging Wp in place of D in Eqs. (1) and (2), we obtain the minimum Wasserstein estimator (MWE) and minimum expected Wasserstein estimator (MEWE) of order p, denoted ??n and ??n,m respectively. Some properties of the MWE have been studied in Bassetti et al. (2006), for well-specified models and i.i.d. data; we propose new results in Section 4. Intuitively, under some conditions we can expect ??n to converge to ??, in the sense that Wp(??n,??) ? 0. Consequently, the minimum of ?  ? Wp(??n,??) might converge to the minimum of ?  ? Wp(??,??), denoted by ??, assuming its existence and unicity. In the well-specified case, ?? coincides with the data-generating parameter. In the misspecified case, ?? is typically different from the limit of the maximum likelihood estimator (MLE), which is the minimizer of KL(??|??).

@@

\subsection{Point estimates}

%\kcnote{connect to MWE and MEWE here?}

When $\rho$ is the Wasserstein distance, $\bftheta^*$ is refered to as the minimum Wasserstein estimator. When the model is well specified, $\bftheta^*$ coincides with the true data-generating parameter; however, if the model is misspecified, $\bftheta^*$ is typically different from the maximum likelihood estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx | \bftheta)$~\citep{bernton2017inference}. In order to more effectively target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially}, Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial}, $\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.  Each of these assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model. 
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of a rejection sampler in an ABC-like setup (as opposed to variational inference).  They emphasize that this approach allows one to eliminate the summary statistics typically used in ABC and calculate the Wasserstein distance explicitly without use Kantorovich-Rubinstein duality and the critic $d$. For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection of the data onto Hilbert space-filling curves. Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality. They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$. This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.
\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL. We also don't specify a prior b/c it's like implicit empirical bayes. They also don't specify a recognition model to latents. }

@@

As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
intimately tied to a class of algorithms that can be framed as density
estimation-by-comparison. In most cases, these inference algorithms are
formulated as an iterative two-step process where the model distribution is
first compared to the true data distribution and then updated to make it more
comparable to the latter. Relevant work in this direction includes \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.

@@

In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients with respect to $\bftheta$ through finite differences from multiple forward passes of the simulator with variance reduction strategies based on controlling the source of randomness used for the latent variable $\bfz$.
